---
title: "MovieLens Recommendation System"
author: "Islam Gamal"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
    fig_caption: true
    fig_width: 7
    fig_height: 4.5
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
```

---

# Introduction

## Overview of the MovieLens Dataset

The MovieLens 10M dataset is one of the most widely used benchmarks in recommender systems research. It was collected by GroupLens Research and contains approximately **10 million ratings** applied to around **10,000 movies** by about **72,000 users**. Each record includes a user ID, a movie ID, a rating (on a scale of 0.5 to 5 in increments of 0.5), a timestamp, and movie metadata such as title and genres.

For this project, the dataset was split into two parts:

- **edx**: the main dataset used for training, development, and tuning (~9 million ratings)
- **final_holdout_test**: a held-out test set (~1 million ratings) used *only* for the final evaluation

One rule I was very careful to follow throughout this project: **I never touched `final_holdout_test` until the very last step**. All model selection, hyperparameter tuning, and intermediate evaluation were done using a validation split created from `edx`. This is standard practice to avoid data leakage.

## Project Objective

The goal is to build a movie recommendation system that predicts the rating a particular user would give to a particular movie. Model performance is measured using **Root Mean Squared Error (RMSE)**:

$$\text{RMSE} = \sqrt{\frac{1}{N} \sum_{i=1}^{N} (\hat{y}_i - y_i)^2}$$

A lower RMSE means our predictions are closer to the actual ratings. The target for this project is to achieve **RMSE < 0.86490**.

## Approach Summary

I started simple and built up complexity step by step:

1. **Baseline model** — just predicting the global mean for everyone
2. **Movie effect model** — accounting for the fact that some movies are generally rated higher than others
3. **Movie + User effect model** — also accounting for individual user tendencies
4. **Regularization** — penalizing estimates from movies/users with very few ratings, to reduce overfitting
5. **Final evaluation** — training on full `edx` and evaluating on `final_holdout_test`

---

# Methods / Analysis

## Data Preparation

The dataset was downloaded, parsed, and joined using the code provided by the course. The key steps were:

- Reading the raw `ratings.dat` and `movies.dat` files using `::` as a delimiter
- Converting columns to the correct data types (userId, movieId as integers; rating as numeric)
- Joining ratings with movie metadata on `movieId`
- Splitting into `edx` (90%) and `final_holdout_test` (10%), ensuring users and movies in the test set also appear in training

```{r data_prep, cache=TRUE}
if(!require(tidyverse))  install.packages("tidyverse",  repos = "http://cran.us.r-project.org")
if(!require(caret))      install.packages("caret",      repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")

library(tidyverse)
library(caret)
library(data.table)

# Download dataset
dl <- "ml-10M100K.zip"
if(!file.exists(dl))
  download.file("https://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)

ratings_file <- "ml-10M100K/ratings.dat"
movies_file  <- "ml-10M100K/movies.dat"

if(!file.exists(ratings_file)) unzip(dl, files = "ml-10M100K/ratings.dat")
if(!file.exists(movies_file))  unzip(dl, files = "ml-10M100K/movies.dat")

ratings <- as.data.frame(
  str_split(read_lines(ratings_file), fixed("::"), simplify = TRUE),
  stringsAsFactors = FALSE)
colnames(ratings) <- c('userId', 'movieId', 'rating', 'timestamp')
ratings <- ratings %>%
  mutate(userId    = as.integer(userId),
         movieId   = as.integer(movieId),
         rating    = as.numeric(rating),
         timestamp = as.integer(timestamp))

movies <- as.data.frame(
  str_split(read_lines(movies_file), fixed("::"), simplify = TRUE),
  stringsAsFactors = FALSE)
colnames(movies) <- c('movieId', 'title', 'genres')
movies <- movies %>% mutate(movieId = as.integer(movieId))

movielens <- left_join(ratings, movies, by = "movieId")

# Train / Final holdout split
set.seed(1, sample.kind = "Rounding")
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx   <- movielens[-test_index, ]
temp  <- movielens[ test_index, ]

final_holdout_test <- temp %>%
  semi_join(edx, by = "movieId") %>%
  semi_join(edx, by = "userId")

removed <- anti_join(temp, final_holdout_test)
edx     <- rbind(edx, removed)
rm(dl, ratings, movies, test_index, temp, removed)
```

## Exploratory Data Analysis

Before jumping into modeling, I spent some time looking at the data. This helped me build intuition about what kinds of signals might be useful.

### Dataset Dimensions

```{r eda_dims}
cat("edx dimensions:", nrow(edx), "x", ncol(edx), "\n")
cat("Distinct users:", n_distinct(edx$userId), "\n")
cat("Distinct movies:", n_distinct(edx$movieId), "\n")
cat("Rating range:", min(edx$rating), "to", max(edx$rating), "\n")
```

### Rating Distribution

```{r eda_ratings, fig.cap="Distribution of user ratings in the edx dataset. Whole-number ratings are much more common than half-star ratings."}
edx %>%
  ggplot(aes(x = rating)) +
  geom_histogram(binwidth = 0.5, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Movie Ratings",
       x = "Rating", y = "Count") +
  theme_minimal()
```

One thing I noticed immediately: **whole-number ratings (1, 2, 3, 4, 5) are far more common than half-star ratings** (1.5, 2.5, etc.). This is just how people tend to rate things — they round to whole numbers. Also, ratings of 4 are the most common, which suggests there's a slight positive bias in user ratings overall.

### Ratings per Movie

```{r eda_movies, fig.cap="Number of ratings per movie on a log scale. The distribution is heavily right-skewed — most movies have few ratings, while a handful have thousands."}
edx %>%
  count(movieId) %>%
  ggplot(aes(x = n)) +
  geom_histogram(bins = 50, fill = "tomato", color = "white") +
  scale_x_log10() +
  labs(title = "Ratings per Movie (log scale)",
       x = "Number of Ratings", y = "Number of Movies") +
  theme_minimal()
```

This is really important: **most movies have very few ratings**, while a small number of blockbusters have thousands. A movie with only 1 or 2 ratings can have an extreme (and unreliable) average. This motivated the use of regularization later on.

### Ratings per User

```{r eda_users, fig.cap="Number of ratings per user on a log scale. Some users are very active raters, while many only rated a handful of movies."}
edx %>%
  count(userId) %>%
  ggplot(aes(x = n)) +
  geom_histogram(bins = 50, fill = "darkgreen", color = "white") +
  scale_x_log10() +
  labs(title = "Ratings per User (log scale)",
       x = "Number of Ratings", y = "Number of Users") +
  theme_minimal()
```

Similarly, **user activity is highly variable**. Some users rated thousands of movies, while many rated fewer than 50. Again, this means some user bias estimates will be based on very little data — regularization will help here too.

### Top 10 Most-Rated Movies

```{r eda_top10}
edx %>%
  group_by(movieId, title) %>%
  summarise(n_ratings = n(), avg_rating = round(mean(rating), 2)) %>%
  arrange(desc(n_ratings)) %>%
  head(10) %>%
  select(title, n_ratings, avg_rating) %>%
  knitr::kable(caption = "Top 10 most-rated movies")
```

## Validation Split

Since I'm not allowed to use `final_holdout_test` during development, I created my own internal validation split from `edx`:

```{r val_split}
set.seed(1, sample.kind = "Rounding")
val_index <- createDataPartition(y = edx$rating, times = 1, p = 0.1, list = FALSE)
train_set <- edx[-val_index, ]
val_temp  <- edx[ val_index, ]

val_set <- val_temp %>%
  semi_join(train_set, by = "movieId") %>%
  semi_join(train_set, by = "userId")

removed_val <- anti_join(val_temp, val_set)
train_set   <- rbind(train_set, removed_val)
rm(val_index, val_temp, removed_val)

cat("Training rows:", nrow(train_set), "\n")
cat("Validation rows:", nrow(val_set), "\n")
```

## RMSE Function

```{r rmse_fn}
RMSE <- function(true_ratings, predicted_ratings) {
  sqrt(mean((true_ratings - predicted_ratings)^2))
}
```

## Modeling

### Model 1: Baseline — Global Mean

The simplest model just predicts the global average rating for every user-movie combination. It doesn't know anything about individual movies or users.

$$\hat{Y}_{u,i} = \mu$$

```{r model1}
mu <- mean(train_set$rating)
cat("Global mean:", mu, "\n")

rmse_baseline <- RMSE(val_set$rating, mu)
cat("Model 1 RMSE:", rmse_baseline, "\n")

results_table <- tibble(Model = "1. Baseline (global mean)", RMSE = rmse_baseline)
```

As expected, the RMSE is around 1.06 — not great. But this is our starting point.

### Model 2: Movie Effect

Looking at the data, it's obvious that some movies are genuinely better (or worse) than average. We model this as a **movie bias** $b_i$:

$$\hat{Y}_{u,i} = \mu + b_i$$

where $b_i$ is estimated as the average of (rating − μ) for each movie.

```{r model2}
movie_avgs <- train_set %>%
  group_by(movieId) %>%
  summarise(b_i = mean(rating - mu))

predicted_movie <- val_set %>%
  left_join(movie_avgs, by = "movieId") %>%
  mutate(pred = mu + b_i) %>%
  pull(pred)

rmse_movie <- RMSE(val_set$rating, predicted_movie)
cat("Model 2 RMSE:", rmse_movie, "\n")

results_table <- bind_rows(results_table,
  tibble(Model = "2. Movie Effect", RMSE = rmse_movie))
```

A solid improvement. Adding the movie effect already brings us well below 1.0.

### Model 3: Movie + User Effect

Even after accounting for movie quality, individual users have their own rating tendencies. Some people are harsh critics (always rate low) while others rate generously. We add a **user bias** $b_u$:

$$\hat{Y}_{u,i} = \mu + b_i + b_u$$

where $b_u$ is the average of (rating − μ − $b_i$) for each user.

```{r model3}
user_avgs <- train_set %>%
  left_join(movie_avgs, by = "movieId") %>%
  group_by(userId) %>%
  summarise(b_u = mean(rating - mu - b_i))

predicted_user <- val_set %>%
  left_join(movie_avgs, by = "movieId") %>%
  left_join(user_avgs,  by = "userId") %>%
  mutate(pred = mu + b_i + b_u) %>%
  pull(pred)

rmse_user <- RMSE(val_set$rating, predicted_user)
cat("Model 3 RMSE:", rmse_user, "\n")

results_table <- bind_rows(results_table,
  tibble(Model = "3. Movie + User Effect", RMSE = rmse_user))
```

Another meaningful improvement. We're getting closer to our target.

### Model 4: Regularized Movie + User Effect

Here's the issue with the previous model: imagine a movie that was only rated by a single person who gave it 5 stars. Its $b_i$ would be huge and positive — but that's not a reliable estimate. Same goes for users who only rated one or two movies.

**Regularization** addresses this by shrinking these estimates toward zero. The penalty is controlled by a parameter $\lambda$:

$$\hat{b}_i(\lambda) = \frac{\sum_{u \in \mathcal{U}(i)} (Y_{u,i} - \mu)}{n_i + \lambda}$$

$$\hat{b}_u(\lambda) = \frac{\sum_{i \in \mathcal{I}(u)} (Y_{u,i} - \mu - \hat{b}_i)}{n_u + \lambda}$$

When $n$ is large, $\lambda$ has little effect. When $n$ is small, $\lambda$ pulls the estimate toward zero — exactly what we want.

I tuned $\lambda$ by searching over a grid of values and selecting the one that minimized RMSE on the validation set.

```{r model4_tune, cache=TRUE, fig.cap="RMSE on the validation set as a function of lambda. The dashed red line marks the optimal value."}
lambdas <- seq(0, 10, 0.25)

rmse_by_lambda <- sapply(lambdas, function(l) {
  b_i_reg <- train_set %>%
    group_by(movieId) %>%
    summarise(b_i = sum(rating - mu) / (n() + l))
  
  b_u_reg <- train_set %>%
    left_join(b_i_reg, by = "movieId") %>%
    group_by(userId) %>%
    summarise(b_u = sum(rating - mu - b_i) / (n() + l))
  
  preds <- val_set %>%
    left_join(b_i_reg, by = "movieId") %>%
    left_join(b_u_reg, by = "userId") %>%
    mutate(pred = mu + b_i + b_u) %>%
    pull(pred)
  
  RMSE(val_set$rating, preds)
})

best_lambda <- lambdas[which.min(rmse_by_lambda)]
cat("Best lambda:", best_lambda, "\n")
cat("Best validation RMSE:", min(rmse_by_lambda), "\n")

tibble(lambda = lambdas, rmse = rmse_by_lambda) %>%
  ggplot(aes(x = lambda, y = rmse)) +
  geom_line(color = "steelblue", size = 0.8) +
  geom_vline(xintercept = best_lambda, linetype = "dashed", color = "red") +
  annotate("text", x = best_lambda + 0.3, y = max(rmse_by_lambda) - 0.001,
           label = paste("lambda =", best_lambda), color = "red", hjust = 0) +
  labs(title = "Lambda Tuning: Regularization Parameter vs RMSE",
       x = "Lambda", y = "RMSE (Validation Set)") +
  theme_minimal()

rmse_regularized_val <- min(rmse_by_lambda)

results_table <- bind_rows(results_table,
  tibble(Model = paste0("4. Regularized (lambda=", best_lambda, ") [Validation]"),
         RMSE  = rmse_regularized_val))
```

---

# Results

## Model Comparison on Validation Set

```{r results_table}
results_table %>%
  knitr::kable(digits = 5, caption = "RMSE comparison across models (on validation set)")
```

The regularized model achieves the lowest RMSE on the validation set. Each step in the modeling progression gave a meaningful improvement:

- Going from the baseline to movie effects: large improvement (~0.11 reduction in RMSE)
- Adding user effects: another strong improvement (~0.08 reduction)
- Adding regularization: a smaller but important refinement

## Final Evaluation on `final_holdout_test`

Now — and only now — I trained the final model on the **full `edx` dataset** (not just the training split) and evaluated it on `final_holdout_test`.

```{r final_model}
# Train final model on full edx
mu_final <- mean(edx$rating)

b_i_final <- edx %>%
  group_by(movieId) %>%
  summarise(b_i = sum(rating - mu_final) / (n() + best_lambda))

b_u_final <- edx %>%
  left_join(b_i_final, by = "movieId") %>%
  group_by(userId) %>%
  summarise(b_u = sum(rating - mu_final - b_i) / (n() + best_lambda))

# Final predictions
final_predictions <- final_holdout_test %>%
  left_join(b_i_final, by = "movieId") %>%
  left_join(b_u_final, by = "userId") %>%
  mutate(pred = mu_final + b_i + b_u) %>%
  pull(pred)

rmse_final <- RMSE(final_holdout_test$rating, final_predictions)
cat("==============================================\n")
cat("FINAL RMSE on final_holdout_test:", rmse_final, "\n")
cat("==============================================\n")

results_table <- bind_rows(results_table,
  tibble(Model = paste0("5. FINAL — Holdout Test (lambda=", best_lambda, ")"),
         RMSE  = rmse_final))

results_table %>%
  knitr::kable(digits = 5, caption = "Complete results summary including final holdout evaluation")
```

The final RMSE on `final_holdout_test` confirms that our model generalizes well. Training on the full `edx` dataset (more data than the training split alone) also helped marginally compared to the validation-set estimate.

---

# Conclusion

## Summary

In this project, I built a movie recommendation system using the MovieLens 10M dataset. Starting from a simple global mean baseline, I progressively added movie effects, user effects, and regularization to improve prediction accuracy. The key findings were:

- **Movie effects matter a lot** — some movies are consistently rated higher or lower than the global mean, and capturing this accounts for a large portion of the variance.
- **User effects also matter** — individual rating tendencies add significant predictive power.
- **Regularization is essential** — without it, estimates for movies and users with few ratings become noisy and hurt performance. Tuning lambda via cross-validation is a clean and effective way to control this.

The final regularized model achieved **RMSE < 0.86490** on `final_holdout_test`, meeting the target for the highest grade.

## Limitations

Despite the solid performance, this model has some clear limitations:

1. **No temporal effects**: Ratings change over time (movie tastes evolve, sequels affect perception of originals, etc.). A time-based component — like allowing biases to drift over time — could improve things further.

2. **No genre information**: The dataset includes genre tags, but I didn't use them. Genre-based effects (e.g., action movies tend to be rated lower than dramas) could add another signal.

3. **Sparsity**: The user-movie matrix is extremely sparse (~99.9% empty). More advanced methods like matrix factorization can extract latent factors from this structure more effectively than simple bias terms.

4. **Cold start problem**: For new users or movies with no ratings, we can't estimate reliable bias terms. The model falls back to the global mean in those cases, which may not be accurate.

## Suggestions for Future Work

If I were to continue this project, I'd explore:

- **Matrix Factorization / SVD**: Methods like `recosystem` or funk-SVD decompose the rating matrix into user and item latent factor matrices, which typically outperform additive bias models.
- **Genre effects**: Adding a genre bias term, similar to the movie and user effects.
- **Temporal effects**: Allowing biases to vary over time, as explored by Koren (2009) in the Netflix Prize competition.
- **Ensemble methods**: Combining multiple models (e.g., bias model + matrix factorization) often gives better results than any single approach.

Overall, this was a genuinely interesting project. It showed me how much mileage you can get out of simple, interpretable models when they're built thoughtfully — and it made me appreciate why regularization is such a fundamental tool in machine learning.

---

# References

- Grouplens Research. *MovieLens 10M Dataset*. https://grouplens.org/datasets/movielens/10m/
- Irizarry, R.A. *Introduction to Data Science*. HarvardX / edX.
- Koren, Y. (2009). *Collaborative filtering with temporal dynamics*. KDD '09.
- Tibshirani, R., Hastie, T., Friedman, J. *The Elements of Statistical Learning* (2nd ed.).
